### 2014-08-11 Studies/2011_San_Francisco/Report/SF_Fire.tex
San Francisco: Updates from Chicago report

% Extra packages
\usepackage{xfrac}

@@

\headerC{
{
\flushright{
Kristopher J. Overholt \\
Daniel Madrzykowski \\
Craig G. Weinschenk \\

@@

\headerC{
\flushright{
Kristopher J. Overholt \\
Daniel Madrzykowski \\
Craig G. Weinschenk \\
{\em Fire Research Division \\
Engineering Laboratory} \\

@@

\begin{tabbing}
\hspace{1.5in} \= \\
MW \> Megawatt \\
s \> second \\
min \> minute \\
m \> meter \\
ft \> foot \\
Pa \> Pascal \\
psi \> pound per square inch \\
k \> thermal  conductivity \\
$\rho$ \> density \\
c$_{p}$ \> specific heat \\
\end{tabbing}

@@

\chapter*{\centering Abstract}
\addcontentsline{toc}{chapter}{Abstract}
% The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on XXXX, within a 2 story, single family structure in XXXX, that resulted in the death of XXXX. The inputs for the FDS simulations are documented in this report. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at face height. Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. The captain was overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure but later succumbed to his injuries.

\chapter{Introduction}
Part of the mission of the Fire Research Division at the National Institute of Standards and Technology (NIST) is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of all of these studies has been to improve the safety of fire fighters.

% On November 2, 2012, a fire in a 2~\sfrac{1}{2} story wood frame residential structure claimed the life of a Chicago fire captain. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using Fire Dynamics Simulator and Smokeview to provide insight into the fire development and thermal conditions that might have existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.1.1) simulations.

\chapter{Fire Incident Summary}
\label{fire_sum}
The account of events for this incident follows the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding times are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines are ``RIT Response'' to be one truck company, one battalion chief, one ALS ambulance, and one EMS field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the [interior door opening into the rear foyer on the back side of the house] and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door [that connects the 1\textsuperscript{st} floor rear porch to the rear foyer], preceded to walk in, saw no fire then backed out.

@@

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~(\ref{fig:simp_geom}).
\begin{table}
\centering
\captionof{table}{Abridged NIOSH approximate fire event timeline}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time  &  Fire Behavior / Fireground Operation  \\
{[hh:mm:ss]}   &                                        \\
\midrule
\multirow{2}{*}{17:15:00} & \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}} \\ 
         & \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}    &  \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}} \\
         & \\[.25cm]
\multirow{2}{*}{17:20:00}    & \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}} \\
         & \\[.25cm] 
17:21:00 & E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline. \\[.25cm]
17:21:30 & First floor exterior porch door opened. \\[.25cm]
\multirow{2}{*}{17:23:00}    & \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}} \\
         & \\[.25cm]
17:24:00    & Water applied to attic from the rear exterior position. \\[.25cm]
17:25:00    & SQ5 makes entry through front of structure. \\[.25cm]
17:27:00    & After hearing a Mayday call from SQ5, IC calls a Mayday. \\[.25cm]
17:29:00    & E123 captain carried out of structure. \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

\chapter{Model Description}
\label{model}
Fire Dynamics Simulator is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}.  A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

Input data from various sources must be collected and documented to simulate a fire using FDS or any other fire model. For the simulation results presented in this document, information was obtained from two primary sources. The following information was gathered from the fire scene: the geometry of the building and the compartments being modeled, the size and location of exterior and interior ventilation openings, and documentation of fire damage to the building. The following information was gathered from witnesses, first responders, reports, and recorded media such as fire ground radios or videos: information on the timing of the fire development, the sequence and approximate timing of ventilation openings to the outside, and weather conditions at the time of the fire. The following subsections will describe the inputs that were used to construct the simulation, including any engineering approximations or assumptions that were made.

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2} story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~(\ref{fig:simp_geom}). This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber with 12.7~mm (0.5~in) gypsum board over top~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~(\ref{fig:simp_geom})), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~(\ref{fig:simp_geom})) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~(\ref{fig:simp_geom})). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~(\ref{fig:simp_geom}). The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~(\ref{fig:alpha_ex}) shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~(\ref{fig:first_floor} and \ref{fig:second_floor}), respectively.

@@

%%%Fig.~(\ref{fig:smv_interior}) shows 3D Smokeview renderings of cutaways of the beta and delta sides of the structure interior.
%%%\begin{figure}[!ht]
%%%\begin{tabular*}{\textwidth}{l@{\extracolsep{\fill}}r}
%%%\includegraphics[width=.80\textwidth]{../Figures/interior_beta} \\
%%%\includegraphics[width=.80\textwidth]{../Figures/interior_delta} \\
%%%\end{tabular*}
%%%\caption{3D Smokeview rendering of structure interior looking from the charlie side to both the beta side (top) and to the delta side (bottom)}
%%%\label{fig:smv_interior}
%%%\end{figure}

@@

To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these findings, the FDS model includes two source fire locations: an initial fire in the attic and a secondary fire on the second floor porch landing. The attic and rear porch primarily consisted of wood fuel: the rear porch was enclosed with wood planks covered with foam board and vinyl siding, the roof consisted of asphalt shingles over wood planking, and the attic and porch floors were composed of oriented strand board.

Wood fuel can be represented by the chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with specified yields of carbon monoxide ($y_{\mathrm{CO}}=0.004 \; \mathrm{kg}/\mathrm{kg}$) and soot ($y_{\mathrm{C}}=0.015 \; \mathrm{kg}/\mathrm{kg}$). The product yields are expressed in terms of the amount of carbon monoxide or soot emitted per unit mass of fuel consumed (kg/kg) and can be found in the SFPE Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood combustion can written as:

@@

To estimate the energy release rate per unit area (kW/m$^2$) of wood, work by Babrauskas and Grayson~\cite{babrauskas1990} is used. Experiments were conducted in a cone calorimeter\footnote{The cone calorimeter is an experimental apparatus used to gather data about ignition time, mass loss, combustion products, and heat release rate among other properties associated with burning small samples of materials.} to determine the 5 minute average of HRRPUA\footnote{HRRPUA is the heat release rate per unit area.} (kW/m$^2$) for several different types of wood over a range of radiant heat fluxes (kW/m$^2$). Babrauskas and Grayson~\cite{babrauskas1990} suggest that the heat release rate per unit area for a variety of woods is approximately 50~kW/m$^2$.

To calculate the total fire size, the appropriate burning area (exposed wood area) of the attic and second floor porch enclosure needs to be determined. Based on the floor plan of the structure, the underside of the roof is approximately 134~m$^2$ (1453~ft$^2$) with 7~m$^2$ (75~ft$^2$) walls on the front and rear sides. Additionally, there is 105~m$^2$ (1130~ft$^2$) of floor area in the attic. Examination of post-incident images indicated that the floor immediately over the rear porch burned away. That area was considered to be part of the porch fire. Based on post-incident damage, approximately 50~\% of the remaining floor area and 90~\% of the roof was considered to be fuel for the attic fire. To estimate the total attic fire size, the underside roof area (120~m$^2$), two end walls (14~m$^2$), and the floor area (43~m$^2$ or 463~ft$^2$) were included. Applying an HRRPUA of wood of 50~kW/m$^2$ results in a potential peak fire size of 9~MW. To estimate the size of the porch fire, the four walls of the porch enclosure (42~m$^2$ or 452~ft$^2$) and the floor and ceiling of the porch (24~m$^2$ or 258 ft$^2$) were included, which results in a total porch area of 66~m$^2$ (689~ft$^2$). Using the same HRRPUA of wood of 50~kW/m$^2$, the potential peak fire size on the porch is approximately 3.3~MW. Therefore, the maximum fire size for this structure was estimated to be 12.3~MW. Since the time at which the fire started is not known relative to the notification of a fire or how fast the fire spread, several simplifications were made. The first assumption is that the attic fire was specified to start at the beginning of the simulation and increase to a value of 9~MW in 10~s. The porch fire was specified to start 40~s into the simulation and increase to a value of 3.3~MW in 10~s. The 10-s ramp up time was used to prevent abrupt changes in the velocity at the fire boundary condition in FDS. The second assumption is that the sources of the gas phase wood fuel have constant areas at fixed locations (2 sources for the attic, 1 source for the porch). The following lines define the source fires in the FDS input file:

\begin{lstlisting}
&SURF ID='FIRE', HRRPUA=2250., COLOR='ORANGE', TAU_Q=10 /
&VENT XB=-13.5,-13,1,5,6.16,6.16, SURF_ID='FIRE' / 
&VENT XB=-14.5,-14,1,5,6.16,6.16, SURF_ID='FIRE' / 
&VENT XB=-16.5,-15.5,1,2.5,3.14,3.14, SURF_ID='FIRE', DEVC_ID='SECOND FIRE'/ 
&DEVC XYZ = -17.2,2,0 ID = 'SECOND FIRE', QUANTITY = 'TIME', SETPOINT = 40, INITIAL_STATE = .FALSE. /
\end{lstlisting}
Note that the coordinates here are unique for the vents are specific to the input files used for these simulations. The key point is that they are 2-D planes located at a specific area.

@@

To examine the results of the simulations, it is important to link the timelines from fire scene to the simulation times. Table~\ref{tab:firesim} shows the fireground timeline along with the corresponding simulation times.

>>>>

@@

\headerC{
{
\flushright{
Kristopher J. Overholt \\
Craig G. Weinschenk \\
Daniel Madrzykowski \\

@@

\headerC{
\flushright{
Kristopher J. Overholt \\
Craig G. Weinschenk \\
Daniel Madrzykowski \\
{\em Fire Research Division \\
Engineering Laboratory} \\

@@

\begin{tabbing}
\hspace{1.5in} \= \\
ft \> foot \\
m \> meter \\
min \> minute \\
Pa \> Pascal \\
psi \> pound per square inch \\
s \> second \\
W \> Watt \\
$k$ \> thermal conductivity \\
$\rho$ \> density \\
$c_{p}$ \> specific heat \\
\end{tabbing}

@@

\chapter*{\centering Abstract}
The Fire Dynamics Simulator (FDS) fire model, which is developed and maintained by the National Institute of Standards and Technology (NIST), was used to provide insight into the dynamics of a fire that occurred on XXXX, within a 2~\sfrac{1}{2}~story, single family structure in San Francisco, CA, that resulted in the death of two firefighters. The inputs for the FDS simulations are documented in this report and are based on the fire scenario, including the building geometry, interior furnishings, and ventilation conditions. The fire started in the attic and spread down through an opening into the enclosed porch on the rear of the structure. This resulted in ventilation limited (fuel rich) fire conditions in the attic and rear porch areas. The spread of fire and hot gases into the second floor of the structure was limited by a closed steel-faced, wood-framed door. After exposure to the elevated temperatures and pressure from the fire in the porch area, the wood frame of the door decomposed, the steel faces of the door failed, and the door collapsed inward. The door failure resulted in the establishment of a flow path between the higher pressure and higher temperature conditions in the enclosed porch and the lower pressure and lower temperature conditions in the hallway and kitchen areas of the second floor. The temperature of the gases that flowed into the hallway exceeded 260~$^{\circ}$C (500~$^{\circ}$F) at a height of 1.86~m (6~ft). Unfortunately, a captain and firefighter were advancing a hose line in the hallway near the door at the time of the door failure. It appears that the captain was likely overcome by the rapid change in conditions before he could exit to the safety of the kitchen area. After a Mayday call, the captain was rescued from the structure, but later succumbed to his injuries.

\chapter{Introduction}
Part of the function of the Fire Research Division at NIST is to develop and apply technology, measurements, and standards to improve the understanding of the behavior, prevention, and control of fires to enhance fire fighting operations and equipment, fire suppression, fire investigations, and disaster response. NIST has previously used the Fire Dynamics Simulator (FDS) fire model to provide insight into the fire development and thermal conditions of fires that have resulted in line of duty deaths (LODDs)~\cite{Madrzykowski:1,Iowa,Texas,Bryner:Charleston,barowy:texas}. The objective of these studies has been to improve the safety and effectiveness of firefighters.

On November 2, 2012, a fire in a 2~\sfrac{1}{2}~story wood frame residential structure claimed the life of a captain of the Chicago Fire Department. NIST examined the fire dynamics of this incident at the request of NIOSH and the Chicago Fire Department. Computer simulations of the fire incident were conducted using the NIST Fire Dynamics Simulator~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide}  software to provide insight into the fire development and thermal conditions that likely existed in the residence during the fire. The specific objectives of the simulations detailed in this report are: 
\begin{enumerate}
\item To examine the effect of the flow path (including temperature, pressure, and fire conditions) in this multi-story residential structure using physics-based calculations.
\item To provide visualizations of the fire behavior that are representative of the conditions that members of the Chicago Fire Department likely experienced during the course of their interior operations.
\end{enumerate}
This document describes the input and the results of the FDS (version 6.0.0) simulations. The simulations were developed using a combination of knowledge of the fire scenario and appropriate engineering approximations and assumptions. Analysis of the simulation results focuses on the hazardous conditions that developed following the development of a flow path inside of the structure. This document is organized as follows: Section~\ref{fire_sum} provides a summary of the fire incident, Section~\ref{model} describes the relevant model input parameters and assumptions that were used to develop the simulations, Section~\ref{results} presents the simulation results, and Section~\ref{discussion} discusses the simulation results as they relate to firefighter safety and effectiveness. Appendix~\ref{app_draw} contains dimensioned drawings of the first and second floor of the structure.

\chapter{Summary of the Fire Incident}
\label{fire_sum}
The account of events for this incident was documented by the National Institute for Occupational Safety and Health (NIOSH) Fire Fighter Fatality Investigation and Prevention Program Report \#F2012-28~\cite{NIOSH:Bowyer}. For the purpose of this analysis, the details regarding the timeline are considered to be approximate values. The following narrative of the incident was extracted from the NIOSH incident report~\cite{NIOSH:Bowyer}:
\begin{quote}
On November 2, 2012, a 54-year-old male career captain sustained injuries at a 2-1/2 story apartment building fire then died at the hospital. At 1716 hours, dispatch called a ``Still'' alarm\footnote{The department defines a ``Still'' alarm to be two engine companies, two truck companies, and a battalion chief.} for smoke in the area. Battalion Chief (BC) 19 was the first to leave the station that was just blocks away. He approached the fire structure by driving behind it then around to the front arriving on scene at 1717 hours. He reported a working fire with heavy smoke coming from the rear and front of the structure's attic. Per fire department standard operating procedures, dispatch initiated a RIT response\footnote{The department defines a rapid intervention team (RIT) response to be one truck company, one battalion chief, one advanced life support (ALS) ambulance, and one emergency medical services (EMS) field officer.}. At 1718 hours, E123 arrived on scene and BC19 was on scene and had assumed incident command (IC). The IC spoke with one of the occupants who stated everyone was out. The IC made entry through the front door to the stairwell to survey the interior of the 2\textsuperscript{nd} floor. He noticed only a light haze throughout and glow around the [rear] door to the covered porch. The IC came back to the front door and met the E123 captain and a firefighter. They had stretched a [hose] load which is 100 feet of 2\sfrac{1}{2}-inch, a gated wye, and 150 feet of 1\sfrac{3}{4}-inch hoseline to go to the 2\textsuperscript{nd} floor, which is a standard department hose lay for this type occupancy. At 1720 hours, E49, T33, and TL39 arrived on scene.

The IC made entry to the 1\textsuperscript{st} floor apartment and worked his way to [the rear of the structure]. He opened the [interior door opening into the rear foyer on the back side of the house] and noticed heavy fire in the covered porch and rear stairwell area. At 1721 hours, the [E123 captain] and firefighter were on the 2\textsuperscript{nd} floor where they flaked out, charged, and began advancing the hoseline to the rear door of the apartment. The E49 crew had stretched a 2\sfrac{1}{2}-inch hoseline [to the rear of the structure]. The T33 crew set a ground ladder on [the left side of the structure] and TL39 set the aerial to the roof about a third of the way back on [the left side]. A TL39 firefighter went to [the rear of the structure] to check doors. He first went to the basement door which he was unable to force open. Then, he went to the first floor exterior enclosed porch door which was unlocked and he opened it up. He stated that he noticed fire light-up in the stairwell. He kicked in the locked door [that connects the 1\textsuperscript{st} floor rear porch to the rear foyer], [proceeded] to walk in, saw no fire then backed out.

@@

\noindent Table~\ref{tab:fire_info} shows an overview of the timeline of events. A plan view of the structure and the location where the captain collapsed are shown in Fig.~\ref{fig:simp_geom}, and more details of the fire development are provided in the following sections.

\begin{table}
\centering
\captionof{table}[Abridged NIOSH approximate fire event timeline]{Abridged NIOSH approximate fire event timeline~\cite{NIOSH:Bowyer}}\label{tab:fire_info}
\begin{tabular}{cl}
\toprule[1.5pt]
Incident Time              &  Fire Behavior / Fireground Operation                                                                                                    \\
{[hh:mm:ss]}               &                                                                                                                                          \\
\midrule
\multirow{2}{*}{17:15:00}  &  \multirow{2}{*}{\parbox{10cm} {Reported working fire with heavy smoke coming from the front and rear of the structure's attic.}}        \\ 
                           &                                                                                                                                          \\[.25cm] %blank rows exist for wrapping text. # of blanks is 1 less than number of rows
\multirow{2}{*}{17:16:00}  &   \multirow{2}{*}{\parbox{10cm} {Dispatch for a ``Still'' alarm - two engine companies, two truck companies, and a battalion chief.}}    \\
                           &                                                                                                                                          \\[.25cm]
\multirow{2}{*}{17:20:00}  &  \multirow{2}{*}{\parbox{10cm} {Attic windows break - heavy smoke out of front attic window and heavy smoke and fire out of rear.}}      \\
                           &                                                                                                                                          \\[.25cm] 
17:21:00                   &  E123 captain and firefighter on second floor with 1\sfrac{3}{4}-inch hoseline.                                                          \\[.25cm]
17:21:30                   &  First floor exterior porch door opened.                                                                                                 \\[.25cm]
\multirow{2}{*}{17:23:00}  &  \multirow{2}{*}{\parbox{10cm} {Heavy fire report in rear stairwell and enclosed porch. No response from E123 captain or firefighter.}}  \\
                           &                                                                                                                                          \\[.25cm]
17:24:00                   &   Water applied to attic from the rear exterior position.                                                                                \\[.25cm]
17:25:00                   &   SQ5 makes entry through front of structure.                                                                                            \\[.25cm]
17:27:00                   &   After hearing a Mayday call from SQ5, IC calls a Mayday.                                                                               \\[.25cm]
17:29:00                   &   E123 captain carried out of structure.                                                                                                 \\
\bottomrule[1.25pt]
\end{tabular}\par
\end{table}

\chapter{Model Description}
\label{model}
Fire Dynamics Simulator~\cite{FDS_Users_Guide} is a computational fluid dynamics (CFD) model that solves a form of the Navier-Stokes equations appropriate for low-speed, thermally driven flow with an emphasis on smoke and heat transport from fires.  Within a CFD model, the room or building is divided into small three-dimensional rectangular control volumes or computational cells.  The cells are contained together within one larger volume known as a computational domain.  The CFD model computes the density, velocity, temperature, pressure, and species concentration of the gas in each cell.  Based on the laws of conservation of mass, momentum,  and energy, the model tracks the generation and movement of fire gasses. One of the most important advantages of FDS is that it is  mathematically verified~\cite{FDS_Verification_Guide} and validated against fire test data to ensure that it provides the expected results, given sufficient input data~\cite{FDS_Validation_Guide}. A complete description of the FDS model is provided in the FDS Technical Reference Guide~\cite{FDS_Math_Guide}.

@@

Input data from various sources must be collected and documented to simulate a fire using FDS or any other fire model. For the simulation results presented in this document, information was obtained from two primary sources. The following information was gathered from the fire scene: the geometry of the building and the compartments being modeled, the size and location of exterior and interior ventilation openings, and documentation of fire damage to the building. The following information was gathered from witnesses, first responders, reports, and recorded media such as fire ground radios or videos: information on the timing of the fire development, the sequence and approximate timing of ventilation openings to the outside, and weather conditions at the time of the fire.

The analysis of the simulation results is focused on the interior conditions on the second floor before and after the interior door failure. Based on the goals of the analysis and the information collected, several engineering approximations were made. For example, the fire exhibited signs of under-ventilated conditions when firefighters arrived~\cite{NIOSH:Bowyer}; however, the exact ignition time, fire growth rate, and heat release rate (HRR) of the fire were not known. The ignition time and growth rate were estimated to reduce the amount of required calculation time while adhering to documented observations. The HRR of the fire was estimated using documentation of the potential fuel load within the structure.

In reality, the quantities associated with these input parameters are not fixed values; rather, a model input parameter can be thought of as a point estimate from a distribution of possible input parameters with some associated amount of uncertainty. Any change in an input parameter (such as the HRR) for a given scenario would result in a change in the output quantity (such as the hot gas layer (HGL) temperature). For example, according to the McCaffrey, Quintiere, and Harkleroad~\cite{SFPE:Walton} empirical correlation, the HGL temperature in a well-ventilated compartment fire is proportional to the HRR raised to the two-thirds power. Following this relationship, a 7.5~\% increase in the HRR would result in a 5~\% increase in the HGL temperature~\cite{NUREG_1824_Sup_1}. More detailed discussion on the propagation of parameter uncertainty in fire models is available in a validation study that was sponsored by the U.S. Nuclear Regulatory Commission~\cite{NUREG_1824_Sup_1}. The following subsections describe the inputs that were used to develop the simulation, including any approximations or assumptions that were made.

\section{Geometry}
\label{geom}
The structure involved in this fire incident was a 2~\sfrac{1}{2}~story residential building with balloon frame construction and a gable roof.\footnote{Balloon frame construction is a method of framing where a continuous stud runs from the bottom horizontal member to the top plate. Intermediate floors are built directly off of the studs. One potential issue with balloon frame construction is that without fire stops there is void space for fire to potentially travel between the floors within the wall cavity.} The structure had a 17.15~m (56~ft) by 6.15~m (20~ft) footprint, as shown in Fig.~\ref{fig:simp_geom}. This results in a floor area of approximately 105~m$^2$ (1100~ft$^2$) per floor and a main living area of approximately 110~m$^2$ (2200~ft$^2$). The interior walls were constructed from nominal lumber covered by 12.7~mm (0.5~in) gypsum board~\cite{NIOSH:Bowyer}. Additionally, there was a half-story attic that contained the structure's hot water heater. Three sets of stairs were present in the structure: one front stairwell that connected the first and second levels (Stair 1 in Fig.~\ref{fig:simp_geom}), one rear stairwell that connected the first and second levels (Stair 2 in Fig.~\ref{fig:simp_geom}) and a second rear set of stairs which provided access to the attic (Stair 3 in Fig.~\ref{fig:simp_geom}). Entry to the first floor of the structure can be made either in the front or rear, as shown in Fig.~\ref{fig:simp_geom}.

The fire, which is described in Section~\ref{fire}, started in the attic and spread to the rear porch. Discussion of the fire's impact on the interior conditions is provided in Section~\ref{results}. Figure~\ref{fig:alpha_ex} shows the exterior of the front and rear of the structure after the incident. Note that the majority of the damage occurred in the attic and rear porch on the second floor. Fully dimensioned drawings of the interior of the first and second floors are shown in Figs.~\ref{fig:first_floor} and \ref{fig:second_floor}, respectively.

@@

\clearpage

@@

To estimate the type of fuel, fire size, and structure ventilation, on-scene videos and post-incident reports were used. Based on these sources of information, the FDS model includes two source fire locations: an initial fire in the attic and a secondary fire on the second floor porch landing. The attic and rear porch primarily consisted of wood fuel: the rear porch was enclosed with wood planks covered with foam board and vinyl siding, the roof consisted of asphalt shingles over wood planking, and the attic and porch floors were composed of oriented strand board.

Wood fuel can be represented by the chemical formula, $\C\H_{1.7}\O_{0.74}\N_{0.002}$, with specified yields of carbon monoxide ($y_{\mathrm{CO}}=0.004 \; \mathrm{kg}/\mathrm{kg}$) and soot ($y_{\mathrm{C}}=0.015 \; \mathrm{kg}/\mathrm{kg}$)~\cite{SFPE:Tewarson}. The product yields are expressed in terms of the amount of carbon monoxide or soot emitted per unit mass of fuel consumed (kg/kg) and can be found in the Society of Fire Protection Engineers (SFPE) Handbook~\cite{SFPE:Tewarson}. A balanced chemical reaction for wood combustion can written as:

@@

To estimate the heat release rate per unit area (HRRPUA) of wood, Babrauskas and Grayson~\cite{babrauskas1990} conducted experiments in a cone calorimeter\footnote{The cone calorimeter is an experimental apparatus used to gather data about ignition time, mass loss, combustion products, and heat release rate among other properties associated with burning small samples of materials~\cite{ASTM:E1355}.} to determine the 5-min average of the HRRPUA for several different types of wood over a range of radiant heat fluxes. The results of that study indicate that the HRRPUA for different types of wood is approximately 50~kW/m$^2$.

To calculate the total fire size, the appropriate burning area (exposed wood area) of the attic and second floor porch enclosure needs to be determined. Based on the floor plan of the structure, the underside of the roof was approximately 134~m$^2$ (1453~ft$^2$) with 7~m$^2$ (75~ft$^2$) walls on the front and rear sides. Additionally, there was 105~m$^2$ (1130~ft$^2$) of floor area in the attic. Examination of post-incident images indicated that the floor immediately over the rear porch burned away. That area was considered to be part of the porch fire. Based on post-incident damage, approximately 50~\% of the remaining floor area and 90~\% of the roof was considered to be fuel for the attic fire. To estimate the total attic fire size, the underside roof area (120~m$^2$), two end walls (14~m$^2$), and the floor area (43~m$^2$ or 463~ft$^2$) were included. Applying an HRRPUA of wood of 50~kW/m$^2$ results in a potential peak fire size of 9~MW. To estimate the size of the porch fire, the four walls of the porch enclosure (42~m$^2$ or 452~ft$^2$) and the floor and ceiling of the porch (24~m$^2$ or 258 ft$^2$) were included, which results in a total porch area of 66~m$^2$ (689~ft$^2$). Using the same HRRPUA of wood of 50~kW/m$^2$, the potential peak fire size on the porch was approximately 3.3~MW. Therefore, the maximum fire size for this structure was estimated to be 12.3~MW. Since the time at which the fire started is not known relative to the notification of a fire or how fast the fire spread, several simplifications were made. The first assumption is that the attic fire was specified to start at the beginning of the simulation and increase to a value of 9~MW in 10~s. The porch fire was specified to start 40~s into the simulation and increase to a value of 3.3~MW in 10~s. The 10-s ramp up time was used to prevent abrupt changes in the velocity at the fire boundary condition in FDS. The second assumption is that the sources of the gas phase wood fuel have constant areas at fixed locations (2 sources for the attic, 1 source for the porch). The following lines define the source fires in the FDS input file:

\begin{lstlisting}
&SURF ID='FIRE', HRRPUA=2250., COLOR='ORANGE', TAU_Q=10 /
&VENT XB=-13.5,-13,1,5,6.16,6.16, SURF_ID='FIRE' / 
&VENT XB=-14.5,-14,1,5,6.16,6.16, SURF_ID='FIRE' / 
&VENT XB=-16.5,-15.5,1,2.5,3.14,3.14, SURF_ID='FIRE', DEVC_ID='SECOND FIRE'/ 
&DEVC XYZ = -17.2,2,0, ID='SECOND FIRE', QUANTITY='TIME', SETPOINT=40, INITIAL_STATE=.FALSE. /
\end{lstlisting}
Note that the coordinates here are unique for the vents are specific to the input files used for these simulations. The key point is that they are 2-D planes located at a specific area.

\clearpage

@@

To examine the results of the simulations, it is important to link the timelines from the fire scene to the simulation times. Table~\ref{tab:firesim} shows the fireground timeline~\cite{NIOSH:Bowyer} along with the corresponding simulation times.
